---
title: The North/South Divide
author: James Riley
date: '2024-02-24'
slug: the-north-south-divide
categories:
  - R
tags:
  - geospatial
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE
)

library("tidyverse")

library("tidyverse")
library("sf")

library("tmap")
tmap_mode("view")
theme_set(ggthemes::theme_few())
```

Various Facebook memes were joking about what parts of England count as "The North" so I decided to grab a shapefile for England and GB, and slap a dot on the map.

I *don't* believe that this defines the North/South in England. I think that issue belongs to social geography and history - in place names and some social geostats you can still see the Danelaw/Anglo-Saxon divide. 

Anyway, the centroid of GB is:

```{r}

gb = rnaturalearth::ne_countries(scale = 50) %>% 
  st_as_sf() %>% 
  filter(name == "United Kingdom") %>% 
  st_transform(27700)

 
  
gb %>% 
  st_centroid(of_largest_polygon = TRUE) %>% 
  #largest polygon is GB in this case.
  tm_shape() + tm_dots()
```

Or for England:

```{r}
pt = rnaturalearth::ne_states(geounit = "england") %>% 
  st_union() %>% 
  st_centroid()

pt %>%  
  
  tm_shape() + 
  tm_dots()

```

There's a charming part of Gaiman's American Gods where the central point of the USA was estimated by means of a cardboard cutout and finding the balancing point, and this place ends up as the opposite as a place of power. 

The brevity of this post disguises how long I spent looking for a good outline of GB and England. The Open Geography Portal wanted to test a new design, it returned no results so I went back to old design.

I kept getting really heavy files from OGP - it is far easier to find all English LSOAs with a bunch of social stats attached than it is to just get the damn coastline. Or to get individual nations but not GB. 

I remembered that there was an illustration of projection of the GB coastline in [The Book](https://r.geocompx.org/spatial-class#vector-data) and found the {rnaturalearth} libraries for grabbing political/coastal borders. 

So ultimately, this turns from a vent about gov open data being hampered by the search function to a love-letter to free/open-source publishing, because I had seen someone else do something like what I wanted, and I could look under the hood for how they did it. 

At the end of the day, this is why I keep this up as a Free/Open Source blog. Even the early, gnarly posts have stuff that someone can learn from, even if it's "this person used to be worse than me!". 
