---
title: Open Coffee Map VI
author: James Riley
date: '2023-11-15'
slug: open-coffee-map-vi
categories:
  - R
tags:
  - geospatial
  - coffee
  - Graph Theory
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE
)
library("valhallr")
library("tidyverse")
library("tmap")
library("sf")

theme_set(ggthemes::theme_few())
```

All map projections are lossy, so there's always a choice in what you preserve.

Or to reframe it, what features you draw attention to. 

Mercator projection preserves angles, which is really handy if you're trying
to sail in a straight line across the Atlantic and reach your port. 

I see the handy "you are here" maps around Leeds, with their neat circles
showing "approx 5 minutes walk" and know that isochrones are not neat circles.

But what if they were?

I had to go under the hood of the {valhallr} package again. I might need 
to offer to take over maintenance of the package.

Valhalla server can return the tree it calculated in making an isochrone, so
I tweaked the isochrone function to go to that API endpoint instead. [^1]

[^1]: Editing code that almost does what you want is so much easier than writing from scratch.

```{r}

val_expand_iso = function (from, costing = "pedestrian", contours = c(5, 10, 
                                                     15), metric = "min", min_road_class = "residential", minimum_reachability = 500, 
          hostname = "localhost", port = 8002) 
{
  if (nrow(from) > 1) 
    stop("More than one location supplied. Please supply a one-row input tibble with `lat` and `lon` columns.")
  if (!(("lat" %in% names(from)) & ("lon" %in% names(from)))) 
    stop("From tibble must inclide one column named `lat` and one named `lon`.")
  if (!metric %in% c("min", "km")) 
    stop("Invalid metric. Please use `min` for time in minutes or `km` for distance in kilometres.")
  post_data <- list()
  post_data$locations <- dplyr::select(from, "lat", "lon")
  post_data$costing <- costing
  if (metric == "min") 
    post_data$contours <- tibble::tibble(time = contours)
  if (metric == "km") 
    post_data$contours <- tibble::tibble(distance = contours)
  
  post_data$action = "isochrone"
  post_data$skip_opposites = TRUE
  post_data$expansion_properties = c( "durations")
  post_json <- jsonlite::toJSON(post_data, auto_unbox = TRUE)
  post_json = str_replace(post_json, '"durations"', '["durations"]')
  url <- paste0("http://", hostname, ":", port, "/expansion")
  resp <- httr::POST(url = url, body = post_json, httr::user_agent("https://github.com/chris31415926535/valhallr"))
  if (httr::http_type(resp) != "application/json") 
    stop("API did not return json.", call. = FALSE)
  if (httr::http_error(resp)) {
    message("Error: API call returned error. Returning API response for debugging.")
    return(resp)
  }
  resp_data <-
  httr::content(resp, type = "text", encoding = "UTF-8") %>% 
    geojsonsf::geojson_sf() %>% tibble::as_tibble() %>% 
    sf::st_as_sf()# %>%
  resp_data$costing <- costing
  return(resp_data)
}


```


```{r}

tmap_mode("plot")

limit = 30

origin = tibble(lat = 53.79806, #t'office
                lon = -1.53158)

# yes, my home server is a homage to Gibson & Wargames movie
data = val_expand_iso(origin, contours = limit, hostname = "gibson.local") %>%
  st_cast("LINESTRING")
```

The unmodified/use GB grid like a normal person version of the map follows:

```{r}
data$durations = jsonlite::fromJSON(data$durations[1])

data = data %>%  
  mutate(durations = durations/60) %>%
  filter(durations < limit) 

data = origin %>% 
  st_as_sf(coords = c("lon", "lat"), crs="EPSG:4326") %>% 
  st_cast("LINESTRING") %>% 
  mutate(durations=0) %>% 
  bind_rows(data) %>% 
  st_transform("EPSG:27700") 

data %>% 
  filter(durations>0) %>% 
  tm_shape() + tm_lines(col="durations")

```

And 'reprojected' so that points that are equal time from the origin are equally far from the origin.

```{r}

origin = st_as_sf( tibble(lat = 53.79806,
                          lon = -1.53158), coords=c("lon", "lat"), crs="EPSG:4326") %>%
  st_transform("EPSG:27700")
  

to = data %>% 
  filter(durations>0) %>% 
  mutate(id=row_number()) %>% 
  st_cast("POINT") %>% 
  distinct() %>% 
  mutate(n=n(), .by=id) %>% 
  filter(n>=2) %>%
  mutate(p = row_number())

#make repeated points have the same durations/r
to = st_coordinates(to) %>%
  as_tibble() %>% 
  mutate(p = row_number()) %>% 
  left_join(to, by="p") %>% 
  group_by(X,Y) %>% 
  mutate(durations=mean(durations)) %>%
  ungroup() %>% 
  st_as_sf()

geom = st_geometry(to)
origin = st_geometry(origin)

r = st_distance(st_set_crs(geom - origin, "EPSG:27700"), origin) %>%
  as.vector()


to = 
  st_set_geometry(to, (geom-origin)*to$durations + origin) %>%
  group_by(id) %>% 
  summarise(durations=first(durations), do_union=TRUE) %>% 
  st_set_crs("EPSG:27700") %>%
  st_cast("LINESTRING")



tm_shape(to) + tm_lines(col="durations")


```

This has already taken long enough. I was going to add the coffee shops from OSM
into both maps, but I have a lot going on right now. Consider this a part 1, but
I'm not obliging myself to finish part 2. 

There's a minor kludge in there - it's not clear if duration applies to the start
or end of a road segment. So roads kept ending up disjointed.

To keep roads connected points that have the same coordinates before transformation
need to have the same coordinates post-transformation, so I:

- split roads into their start and end points;
- initially give both points the duration attached to the road segment;
- gave each point the average duration across all its instances;
- merged them back into lines.

