---
title: Chaos Butterfly
author: James Riley
date: '2019-10-01'
slug: chaos-butterfly
categories:
  - R
tags:
  - simulation
  - visulisation
  - plotly
  - odin
  - chaos
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)
library("tidyverse")
library("lubridate")
library("gganimate")
library("plotly")
set.seed(18159)
```

# Odin

Package [Odin](https://cran.r-project.org/web/packages/odin/index.html) has gone live on CRAN with version 1. It makes some stuff with deSolve a little easier:
- Defining a set of ODEs looks a bit easier
- it's easier to change parameters
- it can write its own C code for when you've got a massive set of ODEs and the efficiency helps.

Also, it's a great punny name.

# Lorentz's Butterfly

Lorentz was trying to study a very simplified weather model with early computers. It's easily described in 3 dimensions with respect to time - [Link to his paper](https://journals.ametsoc.org/doi/abs/10.1175/1520-0469%281963%29020%3C0130%3ADNF%3E2.0.CO%3B2). At least skim it for the pictures, this is 1960's fractal art! The 3 dimensions don't seem to have simple interpretations once they've gone into this form.

It's incredibly sensitive to initial conditions, he had problems trying to resume a simulation from a printout that only had 3 decimals, but the computer held more in memory.

```{r}
library(odin)
library(GGally)

lorentz <- odin({
  deriv(y1) <- sigma * (y2 - y1)
  deriv(y2) <- R * y1 - y2 - y1 * y3
  deriv(y3) <- -b * y3 + y1 * y2
  
  initial(y1) <- start_y_1
  initial(y2) <- start_y_2
  initial(y3) <- start_y_3
  
  start_y_1 <- user(10)
  start_y_2 <- user(1)
  start_y_3 <- user(1)
  
  ## These are the classical parameters:
  sigma <- 10
  R <- 28
  b <-  8 / 3
})

model <- lorentz()

tt <- seq(0, 100, by=1e-3)

model_result <- model$run(tt,method="lsode" ) %>%
  as_tibble() %>%
  mutate_all(as.numeric) 


path <- function(data, mapping){
  ggplot(data=data, mapping=mapping) + 
    geom_path( aes(colour=t)) + 
    scale_color_viridis_c(option = "C") +
    ggthemes::theme_tufte()
}

self <- function(data, mapping){
  ggplot(data=data, mapping=mapping) + geom_point(aes(y=t), size=0.1,alpha=0.1) + coord_flip()
}

model_result %>%
  ggpairs(columns =c("y1","y2","y3"),
          upper=list(continuous=path), 
          diag=list(continuous=self),
          lower=list(continuous=path)) + ggthemes::theme_tufte()
```

It has 3 spatial dimensions, and 1 dimension of time, so let's throw it at more stuff.

```{r}
model_result %>%
  mutate(s=round(t,0)) %>%
  plot_ly(x=~y1,y=~y2,z=~y3, color=~t) %>%
  add_paths()

```

# And if a butterfly flaps its wings

If I give the system a slight nudge in its starting conditions, then it ends up in a totally different bit of its phase space.

```{r}
butterfly <- 1e-6

interfering_butterfly <- lorentz(start_y_1 = 10+butterfly, start_y_2 = 1+butterfly, start_y_3 = 1+butterfly) 

interfering_butterfly <- interfering_butterfly$run(tt)%>%
  as_tibble() %>%
  mutate_all(as.numeric) %>%
  mutate(model="butterfly")

model_result <- mutate(model_result, model="original")

path <- function(data, mapping){
  ggplot(data=data, mapping=mapping) + 
    geom_path( aes(colour=model)) + 
    ggthemes::theme_tufte()
}

self <- function(data, mapping){
  ggplot(data=data, mapping=mapping) + geom_point(aes(y=t, colour=model), size=0.1,alpha=0.1) + coord_flip()
}

bind_rows(model_result, interfering_butterfly) %>%
  filter(between(t,90, 100)) %>%
  ggpairs(columns =c("y1","y2","y3"),
          upper=list(continuous=path), 
          diag=list(continuous=self),
          lower=list(continuous=path)) + ggthemes::theme_tufte()

```


I think the original story involves a printout to 3 decimals, so it's nice to see them disagree quickly for a difference in 6 decimals.
```{r}
bind_rows(model_result, interfering_butterfly) %>%
  plot_ly(x=~y1,y=~y2,z=~y3, split=~model) %>%
  add_paths()
```

# Bonus content - double pendulum

If I'm doing ODEs and chaos, I need a double pendulum.

With thanks to [Schocastics](http://blog.schochastics.net/post/beautiful-chaos-the-double-pendulum/) for making the ODEs for the double pendulum make sense.

```{r}
time <- 10 #seconds
framerate <- 40 #fps

nframes <- time*framerate

double_pendulum <- odin({
  G  <-  9.807  # acceleration due to gravity, in m/s^2
  L1 <-  user(1.0)    # length of pendulum 1 (m)
  L2 <-  user(1.0)    # length of pendulum 2 (m)
  M1 <-  user(1.0)    # mass of pendulum 1 (kg)
  M2 <-  user(1.0)    # mass of pendulum 2 (kg)
  
  pi <- 3.1415926535897
  

  
  # initial conditions
  th1 <-  user(20.0)  # initial angle theta of pendulum 1 (degree)
  w1  <-  user(0.0)   # initial angular velocity of pendulum 1 (degrees per second)
  th2 <-  user(180.0) # initial angle theta of pendulum 2 (degree)
  w2  <-  user(0.0)   # initial angular velocity of pendulum 2 (degrees per second)

  initial(theta1) <- th1*pi/180
  initial(y1) <- w1*pi/180
  initial(theta2) <- th2*pi/180
  initial(y2) <- w2*pi/180
  
  
  deriv(theta1) <-  y1
  
  del_ <-  theta2 - theta1
  den1 <-  (M1 + M2)*L1 - M2*L1*cos(del_)*cos(del_)
  deriv(y1)  <-  (M2*L1*y1*theta2*sin(del_)*cos(del_) +
               M2*G*sin(theta2)*cos(del_) +
               M2*L2*y2*y2*sin(del_) -
               (M1 + M2)*G*sin(theta1))/den1
  
  deriv(theta2) <-  y2
  
  den2 <-  (L2/L1)*den1
  deriv(y2)  <-  (-M2*L2*y2*y2*sin(del_)*cos(del_) +
               (M1 + M2)*G*sin(theta1)*cos(del_) -
               (M1 + M2)*L1*y1*y1*sin(del_) -
               (M1 + M2)*G*sin(theta2))/den2
  
})

L1 <- 1
L2 <- 1/2
M1 <- 1
M2 <- 3

double_pen <- double_pendulum(L1=L1, L2=L2, M1 = M1,M2=M2,
                              th1=90,th2=90)

tt <- seq(0, time, length.out = nframes )

double_pen <- double_pen$run(tt) %>%
  as_tibble() %>%
  mutate_all(as.numeric) %>%
  mutate(x1 = L1 * sin(theta1), y1 = -L1 * cos(theta1)) %>%
  mutate(x2 = x1 + L2*sin(theta2), y2 = y1  -L2*cos(theta2)) 

x <- double_pen %>%
  select(t, x1, x2) %>%
  mutate(x0=0)  %>%
  pivot_longer(starts_with("x"), names_prefix = "^.", values_to="x")  

y <- double_pen %>%
  select(t, y1, y2) %>%
  mutate(y0=0)  %>%
  pivot_longer(starts_with("y"), names_prefix = "^.", values_to="y")

p <- left_join(x, y) %>%
  group_by(t) %>%
  arrange(name) %>%
  mutate(size = c(0,M1,M2)) %>%
  ungroup() %>%
  ggplot(aes(x=x,y=y)) + geom_path() + geom_point(aes(size=size)) + transition_manual(frames=t) + ggraph::theme_graph() + theme(legend.position = "none")

animate(p, nframes=nframes, fps = framerate)

```

  