---
title: What have the Vikings ever done for us?
author: James Riley
date: "`r Sys.Date()`"
slug: what-have-the-vikings-ever-done-for-us
categories:
  - R
tags:
  - geospatial
output:
  blogdown::html_page:
    dev: "svg"
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(lubridate)

library(sf)
library(tmap)

tmap_mode("view")

```

A friend mentioned that "Wensleydale" breaks down into "Wens" (Woden/Wednesday) "ley" (clearing) "dale" (field), or "Odin's field in a woodland clearing". Later, I remembered a lesson in school about elements of place names, and how the [Danelaw](https://en.wikipedia.org/wiki/Danelaw) was more likely to have Viking names than the South. 

We didn't have computers powerful enough to handle the geographic data ONS put out now when I was in school, so I've come back to look at the hypothesis that there are more Old Norse (ON) place names in what was the Danelaw. 

I've tried, unsuccessfully, to get the geography and name for low-level regions from the OpenStreetMap. Querying GB/England all at once makes the server go "nah, that's too big a request", while downloading the XML files means getting into the guts of XML parsing. 

The nearest approximation to what I wanted on ONS was "parishes & non-civil parished areas", but I then lose bigger areas like "Harrogate". 

```{r}
parishes <- st_read("~/bin/R/viking place names/Parishes_and_Non_Civil_Parished_Areas__December_2017__Boundaries_in_England_and_Wales.shp") %>% #File too big for GitHub
  rename(name = parncp17nm) %>%
  select(name) %>%
  mutate(name = str_to_lower(name)) %>% 
  st_centroid()

```


I then went looking for bits of place names that were ON or otherwise, and stumbled into [Wikipedia](https://en.wikipedia.org/wiki/List_of_generic_forms_in_place_names_in_Ireland_and_the_United_Kingdom). That article has plenty of ambiguous forms - "holm" is listed as Old English (OE)/ON. Although OE is Anglo-Saxon, a continental Germanic language - OE & ON are language cousins. [This article](https://en.wikipedia.org/wiki/Toponymy_of_England) has some nice suffixes flagged as OE or ON:

OE:

- -ham
- -ing
- -tun
- -burh
- -ford
- -leah

ON: 

- -by
- -thorp
- -thwait
- -toft

Which makes it look like "Wensleydale" is more OE than ON. 

I'm going to throw a _little_ [regex](https://en.wikipedia.org/wiki/Regular_expression) at this problem, since these are pre-standardised spelling, so "-leah" can also be "-ley". 


```{r}
parishes %>%
  mutate(form = case_when(
    str_detect(name, "(by$|thorp.?$|thwait.?$|toft$)") ~ "ON",
    str_detect(name, "(ham$|ing$|tun$|burh$|ford$|leah$|ley$)") ~ "OE",
    TRUE ~ "OTHER")) %>%
  filter(form != "OTHER") %>%
  tm_shape() + tm_dots(col = "form")
```
What if I'm less strict with suffixes? (Answer, the browser complains about how much data I'm throwing at it, back to static images.)

```{r}
tmap_mode("plot")

parishes %>%
  mutate(form = case_when(
    str_detect(name, "(by|thorp.?|thwait.?|toft)") ~ "ON",
    str_detect(name, "(ham|ing|tun|burh|ford|leah|ley)") ~ "OE",
    TRUE ~ "OTHER")) %>%
  filter(form != "OTHER") %>%
  tm_shape() + tm_dots() + tm_facets("form")

```
I think "-by" shows this pretty well. There's a few false positives, but there's 11,500 parishes in this dataset and it would be a pain to go through al of them by hand.

```{r}
tmap_mode("view")

parishes %>%
  filter(str_detect(name, "by")) %>%
    tm_shape()   +  tm_dots()

```

This makes "Grimsby" "Grim's village", "Grimnir" being "the hooded or masked one (Odin)". The 9 Grim places in this file hardly show a trend:

```{r}
parishes %>%
  filter(str_detect(name, "grim")) %>%
    tm_shape() + tm_dots()
```

I _think_ I'd find more pagan gods in the names of natural features (rivers, hills, etc.) than place names. "Ullswater" _might_ be named for "Ullr" but there's not a Ullswater parish. 

So, Old English place names seem to  be found just about everywhere in England, but Old Norse place names tend to be found in the historic Danelaw. I might come back to looking at the rough borders of the Danelaw in the context of the North-South divide. 