---
title: Garage Gym Comp Results 2022 (final)
author: James Riley
date: '2022-06-08'
slug: ggc-results-2022 
categories:
  - R
tags:
  - GarageGymCompetition
  - fitness
output:
  blogdown::html_page:
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(pins)
library(plotly)
theme_set(ggthemes::theme_few())

```

We have final results for 2022 GGC. [As is tradition, I'm poking the data to see what falls out](https://openor.blog/2021/06/18/garagegym/). 

I did prelim analysis on the prelim results, so I can change the name of the file, tweak a couple of columns, and get the analysis again for near-free. 

```{r}

data = readxl::read_excel("2022-Spring-GGC-Results-Final.xlsx",
                   sheet = 1) %>% 
  janitor::clean_names()  %>% 
  mutate(account = str_c(instagram_username, facebook_profile_url, reddit_username))  


lifts = data %>% 
  select(first_name, sex = athletes_sex,
         age_range, account, 
         body_weight = athletes_body_weight_in_kg,
         total = athletes_total_in_kg, squat = athlete_s_squat_in_kg,
         bench = athlete_s_bench_in_kg, deadlift = athlete_s_deadlift_in_kg) %>% 
  pivot_longer(total:deadlift) 
```


```{r}
lifts %>% 
  filter(sex == "Male") %>% 
  ggplot(aes(x=value)) +
    geom_density() + 
    facet_wrap("name", scales = "free") +
    labs(title = "Male GGC entries")
```


```{r}
lifts %>% 
  filter(sex == "Female") %>%
  ggplot(aes(x=value)) +
    geom_density() + 
    facet_wrap("name", scales = "free") +
    labs(title = "Female GGC entries")
```

Again the field improves :D I'm so proud of us all. 

Again, Non-Binary folk - sorry, I can't really do stats on 2 points. [^1]

[^1]: For stats purposes I put "male" despite identifying as genderNULL. However, that's my choice and I wouldn't impose it on anyone. 

Pulling the histogram from last year, can I make it work?

```{r}
plotly_pipe = function(data){
  data %>% 
    filter(kg > 0) %>% 
    mutate(id = paste0(instagram_name, "|", first_name)) %>% 
    mutate(kg = cut_interval(kg, n= n()/10)) %>% # I learned a new function. Mildly hacky way of dealing with 1/4 as many female entries as male.
    select(id, kg) %>% 
    arrange(kg) %>% 
    group_by(kg) %>% 
    mutate(y = row_number()) %>% 
    ungroup() %>% 
    plot_ly( x = ~kg, y = ~y, hovertext = ~id  ) %>% 
      add_markers()
}

data %>% 
  filter(athletes_sex == "Male") %>% 
  select(first_name, instagram_name = account, kg = athletes_total_in_kg) %>% 
  plotly_pipe()
```

```{r}
data %>% 
  filter(athletes_sex == "Female") %>% 
  select(first_name, instagram_name = account, kg = athletes_total_in_kg) %>% 
  plotly_pipe()
```

Yes! Awesomesauce.

Unlike last year, I've some plans to throw these data into Javascript and make these graphs in JS and let people select themselves. 

For the people who competed last year, how've they been doing?


```{r}
data %>% 
  filter(x2021_total_in_kg > 0) %>% 
  ggplot(aes(x=athletes_total_in_kg, x2021_total_in_kg,
             colour=athletes_sex)) +
    geom_point() +
  labs(x = "2021 total (kg)", y = "2022 total (kg)") +
  coord_fixed()

```

That's... hard to read. Staying still is staying on the line $y=x$. However I have relative changes and a dumbbell plot:

```{r}
data %>% 
  arrange(x2021_total_in_kg) %>% 
  filter(x2021_total_in_kg > 0) %>% 
  mutate(y=row_number()) %>% 
  mutate(improved = x2021_to_2022_percent_improvement > 0) %>% 
  ggplot(aes(y=y, 
             x=athletes_total_in_kg, xend=x2021_total_in_kg,
             colour=improved)) +
  ggalt::geom_dumbbell() 
```

My dumbbell plot doesn't distinguish between the end point being to the left or the right of the start point, so I added a flag for "is their 2022 total above or below the 2021 total". 