---
title: "Workflow"
author: "James Riley"
date: '2018-03-26'
---
```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

library("tibble")
library("knitr")
```
This is an ongoing scratchpad of what I have ideas for, what I'm working on, and what I just need to make presentable. There's a massive difference between code I'm happy gives me an answer, and code I'm happy for someone else to see. 
```{r  echo=FALSE}
Ideas = c("training neural nets", "wikidata querires", "importing Google charts", "filesize to book equivalent", "background: graph theory", "background:spherical coordinates", "background:xml")
Working = c("Making table of workflow", "Open Organograms")
Pretty = c("graph theory on road and train networks")
Housekeeping = c("Add CV/publications", "Include bibliography", "Shortcodes", "Tag posts with GORS compitencies")

n_max=max(length(Ideas), length(Working), length(Pretty), length(Housekeeping))

pad_column <- function(column, n_max){
  return(c(column,rep("",n_max-length(column))))
}
Ideas = pad_column(Ideas,n_max)
Working=pad_column(Working,n_max)
Prettify = pad_column(Pretty, n_max)
Housekeeping=pad_column(Housekeeping,n_max) 
t<-tibble(Ideas,Working, Prettify, Housekeeping)
kable(t)
```

This was originally going to be a Trello. I'm probably going to need something that needs less maintenance than a raw markup file, probably something with vectors.

Now there's a mildly ugly code in the background making a tibble with blanks. But at least it's easier to edit. 