---
title: "Open Organograms"
author: "James Riley"
date: '2018-03-26'
slug: open-organograms-1
tags:
- data.gov.uk
- tidyverse
- data.table
categories: R
--- 
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library("data.table")
library("here")
library("ggplot2")
```
##Background

One of the reasons I've founded this site is because I've discovered https://data.gov.uk/ and I have ideas. In particular, I won't discuss outside of work anything I only know through work, so it was nice to see pay data in the public sphere. 

Full background is at http://guidance.data.gov.uk/organogram-data.html but basically, the Prime Minister decided that the public deserve the structure and hierarchy of the Government Departments, and a few agencies. 

All data used in this post was released under the [Open Government Licence](http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/). 

For reference, there's a good table on [Wikipedia](https://en.wikipedia.org/wiki/Civil_Service_(United_Kingdom)#Grading_schemes) explaining what our grading structure is, including some infor the MOD have given comparing grades in Civil and Military Service.

##Processing

I have decided that not all of these posts will be final finished polished pieces but will include some work in progress. 

Naturally, the first dataset I grabbed was that for the Department of Health and Social Care. Immediately there is a problem - the background above says that the PM wanted everyone to release the data 2ce a year, but our last drop was in 2016. Very often in messing around with data you find that it's out of date. The other thing you find is that there are data standards which get widely ignored.

For now I'm interested in finding median salary, and what a department costs per minute/hour. I need the ggplot practise, so I'm doing some column charts of the distributions too.

##DHSC

This went reasonably smoothly, possibly given that I know the grading scheme. Someone along the line had decided that we wanted info on removed Senior Civil Servants, so I had to filter them out. This processing was done on a data.table framework.

```{r}
c <- read.csv(here("static", "data", "Open Government Licence","department-of-health__31-03-2016__dh_010416-DH-Organogram-junior.csv"), stringsAsFactors = FALSE)
d <-read.csv(here("static", "data","Open Government Licence", "department-of-health__31-03-2016__dh_010416-DH-Organogram-senior.csv"), stringsAsFactors = FALSE)

c <- as.data.table(c)
d<- as.data.table(d)
```

```{r}
juniors = c[,.(N=sum(Number.of.Posts.in.FTE),min=min(Payscale.Minimum....),max=max(Payscale.Maximum....)),by=Grade]
seniors = d[!(Name %in% c("N/D","ELIMINIATED", "VACANT")),.(N=sum(FTE),min=max(Actual.Pay.Floor....),max=max(Actual.Pay.Ceiling....)),by=Grade..or.equivalent.]
```

Then I wanted to append one table to the other. For that they need column names that match. Since I want to find a median I need a cumalitive sum, and to find the middle person.

```{r}
combined <- rbindlist(list(juniors,seniors))

combined <- combined[order(min)] 

combined[,cumulative:=cumsum(N)]

total_staff <- combined[,sum(N)]
median_record <- combined[cumulative>total_staff/2][1]
median_salary = median_record[,min]
```

Then total bill over an estimate at total hours worked in a year gives me the total bill for the department per hour and per minute.

```{r}
total_bill_min <- combined[,sum(N*min)]
total_hours <- (365.25 / 7 * 5 - 25 -9) * (37/5)

avg_hourly <- median_salary/total_hours 
avg_minutely <- avg_hourly/60
```

And so I get my outputs, which I've not nicely formatted yet.

```{r}
total_staff

median_salary

total_bill_min

avg_hourly

ggplot(data=combined, aes(Grade, N)) + geom_col()
```

In fact, that's a totally wrong plot! ggplot has decided it knows my labels better than me and thinks it can do better. I could manually set it, but I'm already past bed time on a work night. 