---
title: UCU is still in pay dispute
author: James Riley
date: '2019-11-23'
slug: ucu-is-still-in-pay-dispute
categories:
  - R
tags:
  - inflation
  - simple graphs
  - ggplot
  - plotly
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)
library("tidyverse")
library("plotly")
set.seed(18223)
```


## The Background

In my short time in [UCU](https://www.ucu.org.uk/) we entered a pay dispute that lead to a series of strike actions. The dispute boiled down to "inflation is eroding our wages, at least pay us as much  *in real terms* as we were worth a few years ago". 

I was a little complicated - I was still a member of [NUS](https://www.nus.org.uk) who had condemned the strikes. I joined UCU as a postgraduate tutor, and was automatically in NUS as a postgraduate student. I ended up on picket lines on my non-tutoring days because timetables are random and UCU decided on random 1 days.

Anyway, I left for the Public Sector, but I have mates in UCU. Despite [UCEA](https://www.ucea.ac.uk) promising to resolve the issues and re-open good-faith negotations, UCU have felt it necessary to vote for industrial action.

I've offered to graph the real-terms cuts to University salaries for one of these mates, and since this blog is [Creative Commons Licenced]() anyone else can grab the graphs

## Data

UCEA [already have commissioned an independent analysis of the real-terms pay cuts](https://www.ucea.ac.uk/library/stakeholder-briefings/newjnchespaypointsrealterms.pdf). I'm going with [CPIH](https://www.ons.gov.uk/economy/inflationandpriceindices/methodologies/consumerpriceinflationincludesall3indicescpihcpiandrpiqmi) as it looks at a standard bucket of living costs including housing, and ONS recommend it and I sit next to a Statistician at work.

I plucked the relevant table into a spreadsheet, giving all values in 2008-09 pounds:

```{r}
pay_data <- read_csv(here::here("static/data/UCU/paycuts.csv"))

pound <- function(x) scales::dollar(x, prefix = "£")

pay_data %>%
  mutate(`Spine Point` = as.character(`Spine Point`)) %>%
  mutate_if(is.numeric, pound) %>%
  knitr::kable()

```

As simple line graphs:

```{r}
parse_year <- function(year) str_extract(year, "^[0-9]+") %>% paste0("-01-01") %>% lubridate::ymd()

pay_data %>%
  pivot_longer(-`Spine Point`, names_to = "year", values_to = "pay") %>%
  mutate(`Spine Point` = as_factor(`Spine Point`)) %>%
  mutate(year =parse_year(year)) %>% #Quick function to turn year into something ggplot recognises as a year
  ggplot(aes(x=year,y=pay, colour=`Spine Point`)) + geom_line() + geom_point() +   ggthemes::theme_tufte() +
  scale_y_continuous(labels=pound) + 
  labs(title="All University Pay Bands have seen real cuts")


```

As percentages:

```{r}
pay_data %>%
 mutate(`Spine Point` = as_factor(`Spine Point`)) %>%
  pivot_longer(-`Spine Point`) %>%
  group_by(`Spine Point`) %>%
  mutate(value = value/first(value)-1) %>%
  ungroup() %>%
  mutate(year = parse_year(name)) %>%
  ggplot(aes(x=year, y=value, colour=`Spine Point`)) + geom_line() + scale_y_continuous(labels = scales::percent) + ggthemes::theme_tufte() + ggthemes::scale_color_few() + labs("Almost everyone is 8% worse off")

```

Spine Point 3 saw less cut than the rest, a quick eyeballing suggests that this is because min wage went up, so I'm re-doing the above graph without SP 3 just in case.


```{r}
pay_data %>%
  filter(`Spine Point`!=3) %>%
  mutate(`Spine Point` = as_factor(`Spine Point`)) %>%
  pivot_longer(-`Spine Point`) %>%
  group_by(`Spine Point`) %>%
  mutate(value = value/first(value)-1) %>%
  ungroup() %>%
  mutate(year = parse_year(name)) %>%
  ggplot(aes(x=year, y=value, colour=`Spine Point`)) + geom_line() + scale_y_continuous(labels = scales::percent) + ggthemes::theme_tufte() + ggthemes::scale_color_few() + labs("Almost everyone is 8% worse off")

```

As a table:

```{r}
pay_data %>%
  mutate(`Spine Point` = as_factor(`Spine Point`)) %>%
  rename(base = `2008-9`) %>%
  pivot_longer(-c(`Spine Point`, base)) %>%
  mutate(value = value/base-1) %>%
  select(`Spine Point`, year=name, value) %>%
  mutate(value = scales::percent(value)) %>%
  pivot_wider(names_from = year, values_from = value) %>%
  knitr::kable()

```

As a time comparison - 8% of a 40-hour week is about 3.2 hours, or half a day. "About half a day" holds as long as they're on a 5-day week, which definitely isn't universal in Uni. (If >40 hours then each day is longer, and half a day is longer.)

Or compared against 10 years ago, most folk in Universities are effectively working for free for half of Friday.