---
title: The joy (plot) of lifting heavy things and putting them down
author: ''
date: '2019-07-04'
slug: the-joy-plot-of-lifting-heavy-things-and-putting-them-down
categories: [R]
tags: 
  - tidyverse
  - Open Powerlifting
  - DYEL
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)
library("tidyverse")
library("ggridges")
library("here")

set.seed(18081)

theme <- ggthemes::theme_few()
scale <- ggthemes::scale_colour_few()
```

## 

A shoutout to [Matthew Analytics](https://mathewanalytics.com/2019/07/01/powerlytics-impact-of-age-gender-and-weight-on-total-weight-lifted-in-powerlifting-meets/) (via R-Bloggers) for linking to the [Open Powerlifting Data](https://www.openpowerlifting.org/data). 

I've started looking at distributions of the lifts (squat, bench, deadlift) that have been performed at meets, and already submitted a bug report on the data. (They were slightly overzealous on matching records with the same name and ended up with _strong_ babies.)

I've also been looking for a reason to use [ggridges](https://cran.r-project.org/web/packages/ggridges/vignettes/gallery.html) so there's this, below:

```{r}
powerlifting_data <- read_csv(here("static", "data", "Open Powerlifting", "openpowerlifting-2019-06-19.csv"))
```

```{r}
powerlifting_data %>%
  filter(Sex=="M", !is.na(AgeClass)) %>%
  mutate(AgeClass = replace(AgeClass, AgeClass=="5-12", "05-12")) %>% #Cheap way to make them line up in order
  select(Sex, AgeClass,
         Bench    = Best3BenchKg,
         Squat    = Best3SquatKg,
         Deadlift = Best3DeadliftKg) %>%
  gather(lift, kg, Bench, Squat, Deadlift) %>% # Make long or 'tidy'
  filter(!is.na(kg)) %>%
  filter(kg>0) %>% # Failed lifts are logged as negative
  ggplot(aes(x=kg, y=AgeClass)) + geom_density_ridges2(aes(rel_min_height=0.005)) + facet_wrap(~lift) + theme_ridges()
```

```{r}

powerlifting_data %>%
  filter(Sex=="F", !is.na(AgeClass)) %>%
  mutate(AgeClass = replace(AgeClass, AgeClass=="5-12", "05-12")) %>% #Cheap way to make them line up in order
  select(Sex, AgeClass,
         Bench    = Best3BenchKg,
         Squat    = Best3SquatKg,
         Deadlift = Best3DeadliftKg) %>%
  gather(lift, kg, Bench, Squat, Deadlift) %>% # Make long or 'tidy'
  filter(!is.na(kg)) %>%
  filter(kg>0) %>% # Failed lifts are logged as negative
  ggplot(aes(x=kg, y=AgeClass)) + geom_density_ridges2(aes(rel_min_height=0.005)) + facet_wrap(~lift) + theme_ridges() 
```

Instagram has chosen to remind me that Eddie Hall is defending his 500 kg world record deadlift. Density plots with thresholds don't really handle single-points, but I could look up his age and add a special labelled point for him.

## Does James even lift?

```{r}
comparison_data <- powerlifting_data %>%
  filter(Sex=="M", AgeClass == "24-34") %>%
   select(
         Bench    = Best3BenchKg,
         Squat    = Best3SquatKg,
         Deadlift = Best3DeadliftKg) %>%
  gather(lift, kg, Bench, Squat, Deadlift) %>% # Make long or 'tidy'
  filter(!is.na(kg)) %>%
  filter(kg>0) 

lift_to_percentile <- function(lift_name, weight){
  
comparison_data %>%
  filter(lift==lift_name) %>%
  arrange(kg) %>%
  mutate(row = row_number()) %>%
  mutate(percentage = row/max(row)) %>%
  filter(kg == weight) %>%
  summarise(from = min(percentage) * 100, to=max(percentage)*100)
}

lift_to_percentile("Bench", 80)

```

So somewhere around the 1st percentile for bench :)

```{r}
lift_to_percentile("Deadlift", 145)
```
And about the 1st percentile for deadlift. I see a pattern.

```{r}
lift_to_percentile("Squat", 130)
```
Trend broken! 2nd percentile for squat. 

This discrepancy makes me want to look at ratioes of personal best squats to deadlifts...

## Squat to deadlift ratio

```{r}
powerlifting_data %>%
  group_by(Name) %>%
  summarise(squat_pr = max(Best3SquatKg), deadlift_pr = max(Best3DeadliftKg)) %>%
  mutate(squat_to_deadlift = deadlift_pr/squat_pr) %>%
  filter(squat_to_deadlift > 0) %>%
  ggplot(aes(x=squat_to_deadlift)) + geom_density() + theme_ridges()
```

That's a loooooong tail, let's zoom in.

```{r}
powerlifting_data %>%
  group_by(Name) %>%
  summarise(squat_pr = max(Best3SquatKg), deadlift_pr = max(Best3DeadliftKg)) %>%
  mutate(squat_to_deadlift = deadlift_pr/squat_pr) %>%
  filter(squat_to_deadlift > 0, squat_to_deadlift < 2.5) %>%
  ggplot(aes(x=squat_to_deadlift)) + geom_density() + theme_ridges()
```

I like the little peak at 1.0. 