---
title: 538-Riddler 2020/03/27
author: James Riley
date: '2020-03-28'
slug: 538-riddler-2020-03-27
categories:
  - R
tags:
  - Riddler
  - Monte Carlo
  - Probability
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)
library("tidyverse")
library("janitor")
library("furrr")

plan("multiprocess")
set.seed(18349L)

theme <- ggthemes::theme_few()
```

Sometimes I see a [Riddler Classic](https://fivethirtyeight.com/features/can-you-get-the-gloves-out-of-the-box/) that makes me just want to go attack it with Monte Carlo analysis.

Let's make a die and roll it 6 times, recording the outcome:
```{r}
die <- 1:6

faces <- sample(seq_along(die), length(die), replace=T) # referring to the length of the die vector will help us if we do the extra credit

die <- die[faces]

die
```

That's 1 iteration. Loops will be necessary since I need the 2nd die to produce the 3rd die, so R's clever vectorisation won't help. 

I'd like to track a 'score' - how far we are from the target of all faces equal. Let's say the score is the number of distinct numbers still on the die, and the target is a score of 1.

```{r}

update_die <- function(die){
  
  faces <- sample(seq_along(die), length(die), replace=T) # referring to the length of the die vector will help us if we do the extra credit

  die[faces]
}
update_score <- function(die){
  unique(die) %>% 
  length()
}

die <- 1:6
score <- update_score(die)

while(last(score) >1){
  die <- update_die(die)
  
  score <- c(score, update_score(die))
}

```

Anything we can do 1ce, we can do many times. 

```{r}
simulate_1_game <- function(index, size_of_die){
  die <- seq_len(size_of_die)
  score <- update_score(die)
  
  while(last(score) >1){
    die <- update_die(die)
    
    score <- c(score, update_score(die))
  }
  
  games <- seq_along(score)
  tibble(index=index, score=score, game=games)
}

monte_carlo_d6 <- map_dfr(seq_len(1e4), ~simulate_1_game(.x,6)) 

```

```{r}
monte_carlo_d6 %>% 
    ggplot(aes(x=game, after_stat(density))) + geom_freqpoly(binwidth=1) + facet_wrap(~score, ncol=1, scales = "free_y") +
  theme + scale_y_continuous(labels=scales::percent)

```

```{r}
monte_carlo_d6 %>% 
  tabyl(game, score) %>% 
  pivot_longer(-game, names_to = "score") %>% 

  ggplot(aes(x=game,y=value,fill=score)) + geom_area() + theme + ggthemes::scale_fill_few()

```

```{r}
monte_carlo_d6 %>% 
  group_by(index) %>% 
  summarise(games_until_win = max(game)) %>% 
  ggplot(aes(x=games_until_win, after_stat(density)))  + geom_freqpoly(binwidth = 1) + 
    theme + scale_y_continuous(labels=scales::percent)


```


```{r}
monte_carlo_d6 %>% 
  group_by(index) %>% 
  summarise(games_until_win = max(game)) %>% 
  tabyl(games_until_win) %>% 
  mutate(percent = scales::percent(percent)) %>% 
  knitr::kable()

```

# Other dice

We'd like to keep track of the size of our dice, tweak the simulate function. Just going to track average games until win to save on memory.

```{r}
rm(monte_carlo_d6)


estimate_victory <- function(size_of_die){

  original_die <- seq_len(size_of_die)
  die <- original_die
  
  dice_pool <- sample(original_die, 1000*size_of_die, replace=T)
  
  play_1_game <- function(...){
    n=0L
    score <- 10 
   
    while(last(score) > 1){
      if(length(dice_pool) < size_of_die){
          dice_pool <<- sample(original_die, 100*size_of_die, replace=T)
      }
      
      n <- n+1L
      die <- die[dice_pool[original_die]] 
      dice_pool <<- dice_pool[-original_die] #Pop 1:n from the dice pool.
      score <- c(score, update_score(die))
    }   
    return(n)
  }
  
  map_int(1:1000, play_1_game) %>% 
    mean()
    
}

results <- future_map_dbl(1:100, estimate_victory)
```
Looks pretty much linear, and might end up more linear if I gave it more simulation time.

```{r}
tibble(size_of_die = 1:100, result=results) %>% 
  ggplot(aes(x=size_of_die,y=result)) + theme + geom_line()

```

