---
title: '#GarageGymCompetition Results'
author: James Riley
date: '2020-06-10'
slug: garagegymcompetition-results
categories:
  - R
tags:
  - DYEL
  - GarageGymCompetiton2020
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)
set.seed(as.Date('2020-06-10'))
library(tidyverse)
library(pins)
library(ggalt)
```

I've had a little play with the [\#GarageGymCompetiton 2020 data](https://garagegymcompetition.wordpress.com/2020-results/) 

First, a density plot of the Male lifts (rescaled to kg) with myself highlighted:

```{r}
data <- pin_get("Garagegym-2020")
me <- data %>% 
  filter(`IG Handle`=="Jimr1603") %>% 
  select(1:5) %>% 
  pivot_longer(-1) %>% 
  mutate(`Weight (kg)`=value/2.2)

data %>% 
  filter(Gender=="Male") %>% 
  select(1:5) %>% 
  pivot_longer(-1) %>% 
  mutate(`Weight (kg)` = value/2.2) %>% 
  ggplot(aes(x=`Weight (kg)`)) + geom_density() +
  geom_vline(data=me, mapping=aes(xintercept=`Weight (kg)`))+ facet_wrap(~name) +
  scale_y_continuous(labels=scales::percent) + ggthemes::theme_few() 
```

I'm considering seeing if I can throw that sort of thing at Plotly in Javascript and let people see themselves on that plot. 

Also, I've discovered [dumbbell charts in package ggalt](https://cran.r-project.org/web/packages/ggalt/vignettes/ggalt_examples.html) so here's a comparison between people's totals for this year against last year:

I'm not squishing 80-ish names into the Y-axis there, so I decided to arrange by this year's total and gave people a number instead of a IG handle.

```{r}
data %>% 
  filter(`Last Year's Total`!=0, Total!=0) %>% 
  arrange(Total) %>% 
  mutate(`IG Handle`=seq_along(`IG Handle`)) %>% 
  ggplot(aes(y=`IG Handle`, x=`Last Year's Total`, xend=Total)) + geom_dumbbell() + 
    ggthemes::theme_few() +
    scale_colour_viridis_c(labels=scales::percent, option = "C") 
```

What about last year's:

```{r}
data %>% 
  filter(`Last Year's Total`!=0, Total!=0) %>% 
  arrange(`Last Year's Total`) %>% 
  mutate(`IG Handle`=seq_along(`IG Handle`)) %>% 
  ggplot(aes(y=`IG Handle`, x=`Last Year's Total`, xend=Total)) + geom_dumbbell() + 
    ggthemes::theme_few() +
    scale_colour_viridis_c(labels=scales::percent, option = "C") 
```

The weirdness is where some people's totals went down. I've not found a nice way to show that on the dumbbell plot.