---
title: Open Coffee Map V
author: james riley
date: '`r Sys.Date()`'
slug: Coffee-V
categories:
  - R
tags:
  - OpenStreetMap
  - geospatial
  - coffee
  - Graph theory
output:
  blogdown::html_page:
    dev: "svg"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library("tidyverse")
library("lubridate")
library("tmap")
library("dodgr")
library("sf")

theme_set(ggthemes::theme_tufte())

```

I've previously looked at the problem of "what coffee shops are in walking distance in a lunchtime from the office?" and [the last attempt](https://openor.blog/2020/02/08/open-coffee-map-part-4/)  involved persuading some OpenStreetMap data to become a Graph. 

I've been reading [Geocomputation with R](https://geocompr.robinlovelace.net/), and one of the definite benefits of publishing with {bookdown} is that it stays up to date. [Chapter 12](https://geocompr.robinlovelace.net/transport.html) introduces a few packages that have a few different approaches to OSM data + street networks. 

For this one I noticed that {dodgr} has "isochrones" - points of equal _time_ from an origin. So I've taken 30 minutes walking from the office to generate a filter on coffee shops.

```{r}
bbox <- tibble(x = c(-1.6200,-1.4852), y = c(53.8223,53.7769) ) %>% 
  as.matrix() 

Leeds <- dodgr_streetnet_sc(bbox) # OSM Data grab

Leeds <- Leeds %>% 
  weight_streetnet(wt_profile = "foot") # Turn it into a Graph, assuming WALKING
```

```{r}
border <- dodgr_isochrones(Leeds, 
                           from = "1675168027", # The vertex outside QH
                           tlim = 30*60)  #seconds
  
#Transform the isochrone into a st_polygon:
border_polygon <- bind_rows(border, slice_head(border, n=1)) %>% 
  select(x,y) %>% 
    as.matrix() %>% 
    list() %>% 
    st_polygon()
```


```{r}
tmap_mode("view")

coffee <- osmdata::opq(bbox) %>% 
  osmdata::add_osm_feature(key = "amenity", value = "cafe") %>% 
  osmdata::osmdata_sf()


filter(coffee$osm_points, st_within(coffee$osm_points, border_polygon, sparse = F)) %>% 
  select(name) %>% 
  tm_shape() + tm_dots()
```

Looking at the extreme points I think {dodgr} has a model of a younger and fitter pedestrian than me. There's quite a lot of hill between the office and "waterside cafe". 

`osmpoints` has sometimes returned the four corners of a coffee shop instead of its center point. This seems to happen when its name is not recorded. The tea shop in Kirkgate Market is not also tagged as a cafe - I might need to go into the documentation to see if it needs an update.

I've a different Shiny/Golem project on the go, so plans to turn this into a Shinyapp are on hold, but with {dodgr} it got a lot easier.

If you've read this far, then I'm going to plug [StreetComplete](https://wiki.openstreetmap.org/wiki/StreetComplete). Just as people like you and me ensure that Wikipedia is as good as it can be, anyone (on Android) can install a small app that asks a few questions like "does this cafe have disabled toilets?" or "what is the speed limit on this road?". When these data are good then people can use them for planning - being able to go into town without worrying about where the facilities are, or where the local vegan eatery is. 