---
title: Riddler 2020-05-15
author: James Riley
date: '2020-05-15'
slug: riddler-2020-05-15
categories:
  - R
tags:
  - Riddler
  - plotly
  - DnD
  - Brute Force
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)
set.seed(as.Date('2020-05-15'))
library(tidyverse)
library(plotly)
```

Two things are guaranteed to grab my attention in a [Riddler Column](https://fivethirtyeight.com/features/can-you-find-the-best-dungeons-dragons-strategy/) - Dungeons and Dragons (DnD) and probability spaces small enough to enumerate all combinations. 

The default position is roll 1 20-sided die (d20), and the other scenarios involve 4 d20 rolls, with a probability space of 20^4= `r scales::comma(20^4)`. More than I'm willing to do by hand, but R laughs at how easy this is.

```{r}

die <- 1:20
all_combinations <- list(die_1 = die,die_2=die,die_3=die,die_4=die) %>% 
  cross_df()
```

```{r}
result_table <- all_combinations %>% 
  mutate(advantage_1 = pmax(die_1,die_2), advantage_2 = pmax(die_3,die_4),
         dis_1 = pmin(die_1,die_2), dis_2=pmin(die_3,die_4)) %>% 
  mutate(dis_of_advantage = pmin(advantage_1, advantage_2),
         advantage_of_dis = pmax(dis_1,dis_2)) 
```

Then the average for each is:

```{r}
result_table %>% 
  summarise(default = mean(die_1),
            dis_of_advantage = mean(dis_of_advantage),
            advantage_of_dis = mean(advantage_of_dis)) %>% 
  knitr::kable()
```

So disadvantage of advantage is better than default, which is better than advantage of disadvantage. On average. 

The probability of getting exactly each roll is:

```{r}
result_table %>% 
  select(die_1, dis_of_advantage, advantage_of_dis) %>% 
  pivot_longer(1:3) %>% 
  count(name,value) %>% 
  group_by(name) %>% 
  mutate(p=n/sum(n)) %>% 
  ungroup() %>% 
  ggplot(aes(x=value,y=p,colour=name)) + geom_line() +
  scale_y_continuous(labels=scales::percent) +
  expand_limits(y=0) +
  theme_minimal()

```

Or the probability of getting _at least_ N on the roll is:

```{r}
result_table %>% 
  select(die_1, dis_of_advantage, advantage_of_dis) %>% 
  pivot_longer(1:3) %>% 
  count(name,value) %>% 
  group_by(name) %>% 
  mutate(p=n/sum(n)) %>% 
  arrange(-value) %>% 
  mutate(p=cumsum(p)) %>% 
  ungroup() %>% 
  plot_ly(x=~value,y=~p,split=~name) %>% 
  add_lines() %>% 
          layout(yaxis = list(tickformat = "%"))
  
```