---
title: Open Coffee Map Part 4
author: James Riley
date: '2020-02-08'
slug: open-coffee-map-part-4
categories:
  - R
tags:
  - OpenStreetMap
  - geospatial
  - coffee
  - Graph theory
  - XML
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)


set.seed(18300)

```

# Statement of Problem

I continue to require coffee at lunch time. If I was doing a Multi-Criteria Decision Analysis I'd throw in some requirements:

* Be able to get to the coffee and back within a 30-60 minute lunch.
* Be reasonably priced.
* Taste good.
* Preference for independent coffee shops. Cost/bucks &c. might be acting _legally_ when they minimise their tax bill, but I can take a stand on the _ethics_ of it by preferring local.
* Getting a reasonable number of steps in is a nice-to-have. If it's not raining.

# Previous Solutions

I looked at this with bits of Python towards the start of this blog, and there's been some developments in R packages for spatial work, interacting with the OpenStreetMap API, and network theory/graph theory.

Currently they're not brilliant at talking to each other! But for these purposes it's not _terrible_ in computing time to load the data twice into different libraries to look at it from different angles - once as geography and once as a network connected by "can walk from A-B". 

# Today's Solution

Normally I put my library calls in the hidden bit at the top, but the serious work is being done by the libraries, so they're going in the main post today:

```{r}
library(tidyverse)
library(sf)
library(osmdata)
library(osmar)
library(tidygraph)
library(tmap)

```

First, I grabbed a sensible bounding box and all the cafes inside that box:

```{r}


bbox <- c(-1.5853357315063479,53.78906626347749,-1.513667106628418,53.81142134499526)

cafe <- opq(bbox = bbox) %>%
  add_osm_feature(key = 'amenity', value="cafe") %>%
  osmdata_sf()

tmap_mode("view")
tm_shape(cafe$osm_points) + tm_dots()
```

So far so good, just need to trim this down to what can be feasibly reached in a lunch break.

```{r}

# Grab footpaths
opq(bbox=bbox) %>% 
  add_osm_feature(key="highway") %>%
  osmdata_xml(here::here("static", "data", "OSM", "footpaths.osm"))

# This step is computationally expensive and I didn't want to do it every time something lower down went wrong.
expensive_parse <- xmlParse(here::here("static", "data", "OSM", "footpaths.osm")) %>%
  as_osmar() 

footpaths <- expensive_parse %>%
  as_igraph() %>% 
  as_tbl_graph() %>% # Footpaths are now in graph representation
  activate(nodes) %>%
  arrange(name!="5506378727") %>% # Manual work to make a node near the enterance to work the first node
  mutate(dist = bfs_dist(root=1)) %>% # Every node labelled with its footpath distance
  as_tibble() # We no longer need the representation as a graph.

# We do need the coordinates of each node to find which paths are near which coffee shops.
nodes <- expensive_parse$nodes$attrs %>% 
  st_as_sf(coords = c("lon", "lat")) %>%
  select(id)

# We get a tibble of coordinates and the shortest walking path distance to them.
path_distance <- nodes %>%
  mutate(name = as.character(id)) %>%
  select(-id) %>%
  left_join(footpaths) %>%
  filter(!is.na(dist)) %>%
  st_set_crs(4326)
```


As far as I can tell, `st_nearest_feature` doesn't like me asking "for every coffee shop what's the nearest footpath point", so I've resorted to a `map_dbl`. 

```{r}

cafe_distance <- function(index){
  
  st_nearest_feature( cafe$osm_points[index,] %>% st_transform(27700)
                      , path_distance %>% st_transform(27700))  %>%
    first() %>%
    slice(path_distance, .) %>%
    st_drop_geometry() %>%
    select(dist) %>%
    dplyr::pull()
}

cafes <- nrow(cafe$osm_points)

cafe <- cafe$osm_points %>%
  mutate(dist = map_dbl(seq_len(cafes), cafe_distance))


```


Finally, plot it!

```{r}
cafe %>%
  select(name, everything()) %>% # Makes name the first column, makes name the mouseover text
  tm_shape() + tm_dots( col="dist", palette = "viridis")
  
```

# Next Steps

If I could eliminate the little bits of manual work I put into this, I could roll it into a shiny app, or generate coffee maps for arbitary points. 