---
title: '#GarageGymCompetition'
author: James Riley
date: '2020-01-20'
slug: garagegymcompetition
categories:
  - R
tags:
  - DYEL
  - tidyverse
  - Fitocracy
  - fitness
  - GarageGymCompetiton2020
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)
library("tidyverse")

garagegym  <- read_csv(here::here("static/data/GarageGym/garagegym2019 - Sheet1.csv"))
```




Previously: I found the [OpenPowerlifting dataset and made some comparisons.](https://openor.blog/post/my-fitocracy-data/). Since Sept last year I've put 10kg on my bench press, 10kg on my deadlift and 20kg on my squat. 

There's going to be the 3rd annual [Garage Gym Competiton](https://www.instagram.com/garagegymcompetition/). (Why you ask, Reddit I reply.) Partially so I can feel like I'm not going to collect the wooden spoon I've grabbed the [2019 results](https://graymatterlifting.wordpress.com/garagegymcompetition-2-2019-results/) and thrown them into a spreadsheet. 

There's a couple of bad values, and Gray Matter thinks in lbs so I'll turn his table into kg.

```{r}
James <- tribble(
  ~Handle, ~Squat, ~Bench, ~Deadlift,
  "James", 170, 100, 170
) %>%
  mutate(Total = Squat + Bench + Deadlift) %>%
  pivot_longer(-`Handle`, values_to = "Weight")

garagegym <- garagegym %>%
  filter_if(is.numeric, ~!is.na(.x)) %>%
  rename(Handle = `G Handle`) %>%
  select(`Handle`, Gender, Squat, Bench, Deadlift, Total) %>%
  pivot_longer(-c(Handle, Gender), values_to = "Weight") %>%
  mutate(Weight = Weight / 2.2) 
```

Some messing around with ggplot2:

First, density plotting everyone's total, excluding me since I didn't take part, but including a vertical line where I am.

```{r}
garagegym %>%
  filter(name=="Total") %>%
  ggplot(aes(x=Weight)) + geom_density() + geom_vline(data=filter(James, name=="Total"), aes(xintercept = Weight)) +
  ggthemes::theme_tufte() + 
  labs(
    title="Density plot of all 2019 competitors, JR highlighted",
    x="Weight (kg)"
  )
```


Facet wrapping the above for the other plots

```{r}
garagegym %>%
  ggplot(aes(x=Weight)) + geom_density() + geom_vline(data=(James), aes(xintercept = Weight)) +
  ggthemes::theme_tufte() + facet_wrap(~name, scales = "free") +
  labs(
    title="Density plots of all 2019 competitors, JR highlighted",
    x="Weight (kg)"
  )
  
```

The comp has no age categories, but there is self-submitted Gender, so comparing male with me:

```{r}
garagegym %>%
  filter(Gender=="Male") %>%
  ggplot(aes(x=Weight)) + geom_density() + geom_vline(data=(James), aes(xintercept = Weight)) +
  ggthemes::theme_tufte() + facet_wrap(~name, scales = "free") +
  labs(
    title="Density plots of all male 2019 competitors, JR highlighted",
    x="Weight (kg)"
  )
  

```

Very similar, 163 male competitors to 41 female competitors.

Out of interest, what were the female numbers like:

```{r}

garagegym %>%
  filter(Gender=="Female") %>%
  ggplot(aes(x=Weight)) + geom_density() +
  ggthemes::theme_tufte() + facet_wrap(~name, scales = "free") +
  labs(
    title="Density plots of all female 2019 competitors",
    x="Weight (kg)"
  )
```


Anyway, in this dataset my everything is just below the mode, except squat. What sort of ranking would I be looking at?

(ha, I thought insta handle would be a unique identifier. I was wrong. I want my data wide again and can't do it. Need to reload the data.)

```{r}
James <-pivot_wider(James, names_from = name, values_from = Weight)


garagegym  <- read_csv(here::here("static/data/GarageGym/garagegym2019 - Sheet1.csv")) %>%
  select(Handle = `G Handle`, Squat, Bench, Deadlift, Total) %>%
  filter_if(is.numeric, ~!is.na(.x)) %>%
  mutate_if(is.numeric, ~(.x/2.2))

garagegym <- bind_rows(James, garagegym)
```


Then I've worked out my percentiles if my current 1RMs were in last year's competition.

```{r}
garagegym %>%
  mutate_if(is.numeric, ~percent_rank(.x)) %>%
  filter(Handle=="James") %>%
  mutate_if(is.numeric, scales::percent) %>%
  knitr::kable()
```



