---
title: Space-filling Curves
author: James Riley
date: '`r Sys.Date()`'
slug: space-filling-curves
categories:
  - R
tags:
  - Mathematics
  - Fractals
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(gghilbertstrings)

theme_set(ggthemes::theme_tufte())

```

# Discovering new packages

My news reader is subscribed to the [R-bloggers news-feed](https://www.r-bloggers.com/2021/04/march-2021-top-40-new-cran-packages/), which includes [RStudio's 'Top 40' new CRAN packages of the month](https://rviews.rstudio.com/2021/04/22/march-2021-top-40-new-cran-packages/). Most of them go over my head - "Genomics" is its own category, and I know next to nothing about Genomics at that level.

I do know a couple of things about Hilbert Curves, so [gghilbertstrings](https://cran.r-project.org/web/packages/gghilbertstrings/index.html) caught my eye. 

I've started a few longer posts and parked them since the last post. Sometimes it's good to just go "this 1 graph is pretty".

# What is this fractal?

[XKCD already used it for a visualisation](https://xkcd.com/195/), and [explain XCKD](https://www.explainxkcd.com/wiki/index.php/195:_Map_of_the_Internet) has a decent explanation of IP addresses, but relies on [Wikipedia](https://en.wikipedia.org/wiki/Hilbert_curve) for explaining the space-filling curve.

# Pictures

gganimate is not playing nicely with my Hilbert curves - you can see an animation on the Wiki page above.

I'm going to choose (too many) digits of $\pi$: I've used it for visualising before.

## Why you don't graph $\pi$ in a line:


```{r}

#Turn Pi into a giant number array. 

pi = read_file(here::here("static/data/pi")) %>% 
  str_remove_all(" ") %>% # Copied from internet, it had some spaces.
  str_remove_all("\\.") %>% 
  str_split("")

pi = pi[[1]] %>% 
  as.numeric()


tibble(x = seq_along(pi), pi = pi) %>% 
  ggplot(aes(x=x, y=pi)) + geom_point()

```

All digits get smushed into each other. If I expand:

```{r}
tibble(x = seq_along(pi), pi = pi) %>% 
  filter(x <= 1000) %>% 
  ggplot(aes(x=x, y=pi)) + geom_point()
```

Still blooming useless. But it looks cool. I think there's a sweet spot where it will look like morse code.

## Hilbert Plots

```{r}
tibble(val = seq_along(pi), pi = pi) %>% 
  filter(val <= 1000) %>% 
  mutate(pi = as_factor(pi)) %>% 
  gghilbertplot(val, color = pi, add_curve = TRUE) + scale_color_viridis_d()

```

By using more of the plot, we can see more.

At 100k points, there's serious overplotting and we're into Jackson Pollock territory. 

```{r}

tibble(val = seq_along(pi), pi = pi) %>% 

  mutate(pi = as_factor(pi)) %>% 
  gghilbertplot(val, color = pi, add_curve = TRUE) + scale_color_viridis_d()
```
