---
title: A "truly horrible" pseudorandom number generator
author: James Riley
date: '2024-07-11'
slug: a-truly-horrible-pseudorandom-number-generator
categories:
  - R
tags:
  - data visualisation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE
)

library("tidyverse")

theme_set(ggthemes::theme_few())

library("plotly")
```

I've gone back to [A History of Data Visualization](https://friendly.github.io/HistDataVis/), 
and there's a nice discussion on the [RANDU algorithm](https://en.wikipedia.org/wiki/RANDU).

The short version is that it's defined by the recurrence relation:

$$ V_{j+1} = 65539 V_j \bmod 2^{31}$$

On 1960's hardware it looked ok. Here's the first few terms starting at $V_0 = 1$, rescaled to the unit interval [0,1]:

```{r}
randu <- 1
for (i in 1:10000) {
  n <- (65539 * last(randu)) %% (2^31)
  randu <- append(randu, n)
}

tibble(randu = randu / (2^31)) %>%
  mutate(x = row_number()) %>%
  ggplot(aes(x = x, y = randu)) +
  geom_point()
```

It takes a little time to get going, but once it's going it looks like it's bouncing all over the place. You can do a boxplot:

```{r}
tibble(x = randu / (2^31)) %>%
  ggplot(aes(x = x)) +
  geom_boxplot()
```

And the median is at 0.5, the quartiles are at 0.25 and 0.75, the whiskers go all the way to 0 and 1...


This algorithm stinks, and Knuth has described it as "truly horrible". 

I wasn't great at number theory, but when I saw the algo something went off in my head. 
Maybe [8 perfect riffle shuffles](https://www.youtube.com/watch?v=MtBzp_-h0oQ) as a highly related concept.

Anyway, what required 5-digit dollar hardware in the 1970s is now available in your web browser.

Behold RANDU in two dimensions, plotting $V_n$ against $V_{n+1}$:

```{r}
tibble(x = randu / (2^31)) %>%
  mutate(y = lag(x)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point()
```
  
In 2d it still looks like it works. But in 3d:


```{r}
tibble(x = randu / (2^31)) %>%
  mutate(y = lag(x)) %>%
  mutate(z = lag(y)) %>%
  plot_ly(x = ~x, y = ~y, z = ~z) %>%
  add_markers(size = 0.1)
```

Drag that around a bit. You're looking for something [like this](https://en.wikipedia.org/wiki/RANDU#/media/File:Randu.png).

I'm going to call this a special case of [Schneier's Law](https://www.schneier.com/blog/archives/2011/04/schneiers_law.html) - anyone can construct a pseudorandom number generator that they can't find a fault with. But with enough eyes any bug is shallow.
