---
title: stationary biking with a raspberry pi
author: James Riley
date: '2023-12-23'
slug: stationary-biking-with-a-raspberry-pi
categories:
  - Python
  - R
tags:
  - cycling
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE
)

library("tidyverse")
library("units")
theme_set(ggthemes::theme_few())

```

As well as lifting heavy weights, I cycle for my health. When it's this cold,
wet and windy I slot my bike into a stationary setup. 

I'll do some proper project photos and cross-link to a hardware hacking space, 
but I've connected a cheap bike computer (magnet + reed switch) to a Raspberry Pi,
which has been told to log when this switch closes, once per revolution.

[Github for the code for the bike](https://github.com/jimr1603/stationary-bike)

Anyway, I'm iterating this - test, refine, test.

Let's load the last two rides and fix the current issues.

```{r}
ride_2 = read_csv("https://github.com/jimr1603/stationary-bike/raw/main/2023-12-22%2020_18", col_names="time" ) %>% 
  mutate(time = as.POSIXct(time))
ride_1 = read_csv("https://github.com/jimr1603/stationary-bike/raw/main/2023-12-20%2017_01", col_names = "time") %>% 
  mutate(time = as.POSIXct(time))

```

```{r}
circumference = 2.21 # wheel circumference, in meters.

ride_1 %>% 
  mutate(dt = as.numeric(time-lag(time))) %>% 
  mutate(dx = as_units(circumference, "m")) %>% 
  mutate(dt =as_units(dt, "seconds")) %>% 
  mutate(speed = dx/dt) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  ggplot(aes(x=time,y=speed)) + geom_point()
```

Negative speed?! Let's remove that first.

```{r}

ride_1 %>% 
  mutate(dt = as.numeric(time-lag(time))) %>% 
  mutate(dx = as_units(circumference, "m")) %>% 
  mutate(dt =as_units(dt, "seconds")) %>% 
  mutate(speed = dx/dt) %>% 
  filter(as.numeric(speed)>0) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  ggplot(aes(x=time,y=speed)) + geom_point()
```

2000 miles per hour is so obviously bad, I'm going to try to zoom in. (Skipping several trial-and-error steps in where I choose my cutoff :D )

```{r}
ride_1 %>% 
  mutate(dt = as.numeric(time-lag(time))) %>% 
  mutate(dx = as_units(circumference, "m")) %>% 
  mutate(dt =as_units(dt, "seconds")) %>% 
  mutate(speed = dx/dt) %>% 
  filter(as.numeric(speed)>0) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  filter(speed <= as_units(100, "miles/hour")) %>% 
  ggplot(aes(x=time,y=speed)) + geom_point()
```

So there's a nice gap - under "100 miles/hour" the readings look sensible (except for negative time), and all readings over 100 miles per hour are sensor barf.

```{r}
ride_1 %>% 
  mutate(dt = as.numeric(time-lag(time))) %>% 
  mutate(dx = as_units(circumference, "m")) %>% 
  mutate(dt =as_units(dt, "seconds")) %>% 
  mutate(speed = dx/dt) %>% 
  filter(as.numeric(speed)>0) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  filter(speed >= as_units(100, "miles/hour")) %>% 
  slice_min(speed)
```

It looks like the debounce time of 0.01 seconds is too small, I'll try a debounce time of 0.03 for the next run.

If I manually throw out readings that are less than 0.03 secs from the last reading [^1] do the data look sensible?

[^1]: incidentally fixing negative readings

```{r}
ride_2 = read_csv("https://github.com/jimr1603/stationary-bike/raw/main/2023-12-22%2020_18", col_names="time" ) %>% 
  mutate(dt = time-lag(time)) %>%
  filter(dt > 0.03) %>% 
  mutate(dt = time-lag(time)) %>% 
  mutate(time = as.POSIXct(time))
ride_1 = read_csv("https://github.com/jimr1603/stationary-bike/raw/main/2023-12-20%2017_01", col_names = "time") %>% 
  mutate(dt = time-lag(time)) %>% 
  filter(dt > 0.03) %>% 
  mutate(dt = time-lag(time)) %>% 
  mutate(time = as.POSIXct(time))
```

```{r}
ride_1 %>% 
  mutate(dt = as_units(dt, "seconds")) %>% 
  mutate(dx = as_units(circumference, "m")) %>% 
  mutate(speed = dx/dt) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  ggplot(aes(x=time,y=speed)) + geom_point()

```

[^2]


[^2]: I was listening to [The Witching Hour](https://lesbianbeddeathband.bandcamp.com/album/the-witching-hour) during this ride. I might align speed, heart rate, and track listing later. 


```{r}
ride_2 %>% 
  mutate(dt = as_units(dt, "seconds")) %>% 
  mutate(dx = as_units(circumference, "m")) %>% 
  mutate(speed = dx/dt) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  ggplot(aes(x=time,y=speed)) + geom_point()

```

Even the ride I held back from the first analysis looks good :D

```{r}
bind_rows(
  mutate(ride_1, id=1),
  mutate(ride_2, id=2)
) %>% 
  summarise(distance = circumference*n(), 
            duration = as.numeric(last(time)-first(time)),
            .by="id") %>% 
  mutate(distance = as_units(distance, "m"),
         duration = as_units(duration, "minutes")) %>% 
  mutate(avg_speed = set_units(distance/duration,"miles/hour")) %>% 
  knitr::kable()
```

Eyeballing recent rides, 10 or 11 miles per hour looks about right, give or take some hills. 
I'll keep the wheel circumference of 2.21 m, push a debounce on the switch of 0.03 seconds, and have another ride.


