---
title: "Unit conversion"
author: "James Riley"
date: "`r Sys.Date()`"
slug: unit-conversion
categories: R
tags:
- cycling
- units
- Visualisation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(units)
library(ggforce)
theme_set(ggthemes::theme_few())

```

If I had to condense advice for life/wellbeing/career into a soundbite, I'd go with "be curious". 

I've been comparing cycle routes on [CycleStreets](https://www.cyclestreets.net/help/journey/faq/#calories), Google Maps, and Strava. CycleStreets has a energy cost estimate of the journey, with a little ? link next to it. 

[They outline their estimate of calorie cost of a journey, including their formula.](https://www.cyclestreets.net/help/journey/faq/#calories)

This is a *brilliant* example of the mess of units we use. The formula was written for meters/second, I think of road speed in miles per hour, a lot of the world thinks of kilometres per hour. I think of human energy in kcal, the formula wants to use watts. 

All my route finders give a total ascent/descent stat, but if I want my distances in miles they give it in feet, when I think of ascent in meters. 

## Subscribe to R-bloggers already

I have [R-bloggers](https://www.r-bloggers.com/) in my News Reader. A while ago the authors of the package announced [units](https://cran.r-project.org/web/packages/units/index.html) on R-bloggers. 

I had my formula in science units, my head in a mess of units, but I remembered that I spotted units had been released. 

## Back to energy spent cycling

A first stab at making that formula as a function:

```{r}
power = function(v, 
                 g=9.81, 
                 m=90, 
                 k_1=0.0053, 
                 k_2=0.26,
                 s=0){
  g * m * v * (k_1 + s) + k_2 * v^3
} 

power(5.4)

```

I simplified it a bit by saying v_a = v_g. And we've replicated their example.

If I do the sums, then apply units I can look at a speed-power curve:

```{r}
tibble(speed = seq(0, 11, length.out= 100)) %>% 
  mutate(power = power(speed)) %>% 
  mutate(speed = set_units(speed, "m/s")) %>% 
  mutate(power = set_units(power, "watts")) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% #SI to thinking in road speed
  mutate(power = set_units(power, "kcal/hour")) %>% #SI to human-level energy units
  ggplot(aes(x=speed, y=power)) + geom_line()
```

This is after I tried starting with speed in miles/h and converting to m/s. It didn't like that because the units are incompatible. 

This is great! Sensible errors are better than producing nonsensical output, so let's rewrite the function including a statement that most of these parameters have dimension.


```{r}
power = function(v, 
                 g=9.81, 
                 m=90, 
                 k_1=0.0053, 
                 k_2=0.26,
                 s=0){
  units(v) <- "m/s"
  units(g) <- "m/s^2"
  units(m) <- "kg"
  units(k_1) <- 1
  units(k_2) <- "kg/m"
  units(s) <- 1
  set_units(g * m * v * (k_1 + s) + k_2 * v^3, "watts")
} 

```


Then what I wanted for the above example is:


```{r}
tibble(speed = seq(0, 25, length.out = 100)) %>% #I think this is clearer as 2 lines 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  mutate(power = power(speed)) %>% 
  mutate(power = set_units(power, "kcal/hour")) %>% 
  ggplot(aes(x=speed, y=power)) + geom_line()
```

## Other graphs

I read [Sustainable Energy Without the Hot Air](https://withouthotair.com/), which suggested in general that moving slower is more energy efficient for distance. [^1]

[^1]: That formula sure looks like it'd apply to a motor vehicle, with some different k_1 and k_2.


I scribbled on my notepad:

I have:

- speed $\frac{m}{s}$
- power $\frac{j}{s}$

I want:

- energy per distance $\frac{j}{m} = \frac{j}{s} \div \frac{m}{s} = \frac{\text{power}}{\text{speed}}$

So the dimensional analysis helped me work out what sums I wanted to do. 

```{r}
tibble(speed = seq(0, 25, length.out = 1000)) %>% #I think this is clearer as 2 lines 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  mutate(power = power(speed)) %>% 
  mutate(energy_per_dist = power/speed) %>% 
  mutate(energy_per_dist = set_units(energy_per_dist, "kcal/mile")) %>% 
  ggplot(aes(x=speed, y=energy_per_dist)) + geom_line() +
  ggforce::scale_y_unit(breaks = c(0,5,10,15)) + expand_limits(y=c(0,15))

```

Or using "fat is 9kcal/gram":

```{r}
tibble(speed = seq(0, 20, length.out = 1000)) %>% 
  mutate(speed = set_units(speed, "miles/hour")) %>% 
  mutate(power = power(speed)) %>% 
  mutate(energy_per_dist = power/speed) %>% 
  mutate(energy_per_dist = set_units(energy_per_dist, "kcal/mile")) %>% 
  mutate(fat_burned = energy_per_dist / as_units(9, "kcal/g")) %>% 
  mutate(fat_burned = set_units(fat_burned, "g/Mm")) %>% 
  ggplot(aes(x=speed, y=fat_burned)) + geom_line() +
  expand_limits(y=0)

```

"Mm" being "mega-meters", or 1000 km. Somewhere in the region of 40 office commutes = 1Mm, for me. I suppose this reinforces the truism that one does cardio to improve their cardiovascular health, not for the energy deficit! 

## Conclusion

- The {units} package can simplify a bunch of analysis, directly by handling conversion, and indirectly by pointing out that you're trying to add meters and seconds. 
- Cycling can be part of my weight loss, but not the whole story. 