---
title: "Two letter English words"
author: "james riley"
date: "`r Sys.Date()`"
slug: []
categories: R
tags: visulisation
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(plotly)

theme_set(ggthemes::theme_tufte())
```
A short one today as I just wanted to see these two graphs.

[The Oatmeal has a free game](http://theoatmeal.com/blog/free_game) with his usual humour. It's built on making words from the given letters, and one level only lets you play two letter words. 

I found an English dictionary-file at [github](https://raw.githubusercontent.com/dwyl/english-words/master/words_alpha.txt), so let's load it and keep the two letter words.

```{r}
two_letter <- read_lines("https://raw.githubusercontent.com/dwyl/english-words/master/words_alpha.txt") 
two_letter <- tibble(word = two_letter) %>% 
  filter(str_length(word) == 2)
```

There are $26^2=674$ possible 2-letter words, but only 427 of them are realised. "ta" is a word, "tf" is not. 

A little tidymagic will visualise this:

```{r}
two_letter %>% 
  mutate(letter_1 = str_extract(word, "^.")) %>% 
  mutate(letter_2 = str_extract(word, ".$")) %>% 
  ggplot(aes(x=letter_1, y=letter_2)) + geom_tile() + coord_fixed() +
    labs(title = "Two-letter English Words")

```



Or some plotly magic:


```{r}
two_letter %>% 
  mutate(letter_1 = str_extract(word, "^.")) %>% 
  mutate(letter_2 = str_extract(word, ".$")) %>% 
  plot_ly(x=~letter_1, y=~letter_2) %>% 
    add_markers(text=~word, hoverinfo="text") 
```



