---
title: Pinning OPL Data
author: James Riley
date: "`r Sys.Date()`"
slug: pinning-opl-data
categories:
  - R
tags:
  - DYEL
  - fitness
  - Open Powerlifting
  - package development
output:
  blogdown::html_page:
    dev: svg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)


theme_set(ggthemes::theme_tufte())

```


RStudio anounced [pins 1.0.0](https://cran.r-project.org/web/packages/pins/vignettes/pins.html). I like the changes, but it's taking a bit of getting used to.

*Find data and put it somewhere sensible* is a good first step in a project. In the old API it was difficult to distinguish between "here is a dataset I have that will never change, put it somewhere sensible" and "I would like the latest data from this server, but I'd like to cache it so I don't hit the server every time I render this report". 

For example, I've handwaved "2020 is a weird year for Powerlifting so I'm excluding it" in a few posts so I don't have to update my OpenPowerlifting pin. 

If this was as simple as "I pointed `board_url` at the file" then I wouldn't be posting. The file is a zip. {readr} can read a csv inside a zip but I can only find it working when the zip consists of only a csv. 

So I hacked together a simple script that pointed `board_url` at the zip, extracted the csv and pinned it. 

I don't want the same snippet at the start of any OPL analysis, so I took a few additional steps to make it a package - [openpoweRlifting](https://github.com/jimr1603/openpoweRlifting). [^1]

[^1]: Look, there's two hard problems in CompSci - naming things and cache invalidation. And I've delegated the 2nd problem to {pins}.

It's a minimal product at this point, you can throw a PR up for that sweet [Hacktoberfest](https://hacktoberfest.digitalocean.com/) swag if you want.

## RStudio Cloud

RStudio server on my home box is out of date, and I've given that box a little bit too much work. I'm testing RStudio Cloud. So far my only problem is that I want certain settings by default in a new workspace. It's probably here somewhere. 

As this blog works via GitHub I only needed to clone the project to work over here. 

Similarly, I have a fresh environment to test my new pinboard!

## Pinning the OPL

```{r}
library("pins")
remotes::install_github("jimr1603/openpoweRlifting")
library("openpoweRlifting")

board = board_folder("~/pins/")

#pin_opl(board)

opl = pin_read(board, "opl-ipf")  %>%
  filter(equipment == "Raw") %>% #free up a little RAM
  select(sex, age_class, bodyweight_kg, weight_class_kg, best3squat_kg,
         best3bench_kg, best3deadlift_kg) %>% #free up more RAM
  filter((sex=="M" & weight_class_kg %in% c('59', '66', '74', '83', '93', '105', '120', '120+')) |
          (sex=="F" & weight_class_kg %in% c('47', '52', '57', '63', '69', '76', '84', '84+'))) # filter for current wt classes. 
```


There's two expensive computations there - downloading the zip and extracting the zip. {pins} ensures that the zip doesn't get downloaded again unnecessarily. I may or may not get around to fixing the other one. (Looks at so many unfinished projects on GH.)


## Analysis

A recent episode of [Iron Culture Podcast](https://www.youtube.com/channel/UCEGGAs257niPVJ5BvXymVLQ) floated the suggestion that some weight categories suit different lifts more. So I'm going to do a simple analysis on max lifts by weight class & sex. 


```{r}
opl %>%
  mutate(id = row_number()) %>% #cheaper on RAM than name strings
  pivot_longer(c(best3squat_kg, best3bench_kg, best3deadlift_kg)) %>%
  filter(!is.na(value)) %>%
  group_by(name, weight_class_kg, sex) %>%
  summarise(max_lift = max(value), bodyweight = mean(bodyweight_kg, na.rm=T)) %>% #Hack to make the graph look right
  ungroup() %>%

  rename(lift=name) %>%
  ggplot(aes(x=bodyweight, y=max_lift, colour=lift)) + geom_point() + facet_wrap("sex") + ggthemes::scale_colour_few()

```
While the general trend is increasing for all 3 lifts X 2 sexes [^2], there are a few cases where the lighter weight class is doing better than the next class up, on absolute numbers. 


[^2]: A very small number of lifters are recorded as Mx. The [2021 Rulebook](https://www.powerlifting.sport/fileadmin/ipf/data/rules/technical-rules/english/IPF_Technical_Rules_Book_2021.pdf) only has weight classes for M/F and otherwise only references Men and Women. Today the easiest thing to do is filter for sex %in% c(M, F), but digging further into how this data happened is on the ideas pile.

## Ideas Pile

Somewhere in-and-among all this I found a couple of bugs in the package. (There's 1 function, and I've fixed 2 bugs already...) and updated this post. 

The graph of absolute top lifts by weight category is interesting, but it'd be nice to come back to this and look at [Allometric Scalling](https://www.strongerbyscience.com/whos-the-most-impressive-powerlifter/). 
The free tier of RS Cloud is limited to 1GB of RAM. This was not a good dataset to work with. I'm going to create an issue where someone can pin the dataset as a SQLite Database to make analysis easier on the free machine. 